---
title: Measuring effect through experimenting
author: "András Bérczi"
date: "October 9, 2019"
output: html_document
# runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '..')
```

# 1. Calculate the Minimum Detectable Effect for a sample of 1000 contacts, when the base rate is 10%. Use the `calculateMinimumDetectableEffect` function!

```{r function to calculate the Minimum Detectable Effect}
calculateMinimumDetectableEffect <- function(p1, n, prop_of_treatment = 0.5, alpha = 0.05, beta = 0.2) {
    M <- qt(1-alpha/2, n - 1) + qt(1-beta, n - 1)
    variance <- p1 * (1 - p1)
    weight <- prop_of_treatment * (1 - prop_of_treatment) * n
    M * sqrt(variance / weight)
}
```

# 2. Calculate the effect for the whole period of the [provided example experiment](./experiment_result.csv)

# 3. Plot the results

# 4. Calculate the effect week-by-week

# 5. Plot the results

# 6. How would you plot uncertainty? Try to do it for both plots!