# ggplot2 prerequisites

> + Learn to finish the below tasks by understanding the [basics](#basics) and then using the first 3 links from the [resources section](#resources).  
> + Learn more advanced features (optional) by using the last link from [resources](#resources).

## Installing dependencies
```{r, warning=FALSE}
    # install.packages("ggplot2")
    library(ggplot2)
```

## Data

We'll use the *diamonds* data frame (`?ggplot2::diamonds`). Its first 6 rows look like the this:

```{r, echo=FALSE}
    head(diamonds)
```
## Basic concepts of *Grammar of Graphics* {#basics}

  + "gg" in ggplot means: Grammar of Graphics  
  + Grammar of Graphics is about mapping data points to aesthetic attributes (colour, shape, size) of geometric objects (points, lines, bars)  
  + Thus, ggplot works with data frames and not individual vectors and maps data from the data frame to parts of the plot (aesthetics)  
  + You can keep enhancing the plot by adding more layers  
  + basic syntax: `ggplot(data = NULL, mapping = aes()) + geom_xyz()`  
      * where you provide the data frame to the `data` argument  
      * and tell the `mapping` argument how to map columns of your data frame to certain parts of your plot  
      * Note that the `mapping` argument's value should be wrapped in the `aes()` (aesthetics) function. This tells the ggplot function which parts of the plot should be the mapped to the data frame. Plot aesthetics defined outside of `aes()` will be constants. (Take a look at the examples and come back here again after to make sure you understood.)  
  + type `?ggplot` to your console to learn more about the arguments  


## Basic plot

```{r}
    ggplot(diamonds, aes(x=carat, y=price, color=cut)) +
        geom_point()
```

## Layers (you can do the same thing in different ways)

```{r, eval=FALSE}
    ggplot(diamonds, aes(x=carat, y=price, color=cut)) +
        geom_point() +
        geom_smooth()
```

_and_

```{r, eval=FALSE}
    ggplot(diamonds) +
        geom_point(aes(x=carat, y=price, color=cut)) +
        geom_smooth(aes(x=carat, y=price, color=cut))
```

are the same.

```{r, echo=FALSE}
    ggplot(diamonds) +
        geom_point(aes(x=carat, y=price, color=cut)) +
        geom_smooth(aes(x=carat, y=price, color=cut))
```

## Titles and labels

## More advanced features

## Basic style guide

## Resources {#resources}

  + [Selva Prabhakaran - The Complete ggplot2 Tutorial](http://r-statistics.co/Complete-Ggplot2-Tutorial-Part1-With-R-Code.html)
  + [Quick ref.](http://r-statistics.co/ggplot2-cheatsheet.html)
  + [Cheat sheet](http://r-statistics.co/ggplot2-cheatsheet.html)

  + [Hadley Wickham - ggplot2 - Elegant graphics for data analysis](http://moderngraphics11.pbworks.com/f/ggplot2-Book09hWickham.pdf)


