---
title: "Eltecon Data Science Course"
subtitle: "Introduction - Home work"
output:
  pdf_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Send your solution in a file named `intro_hw_{your_name}.R` to eltecon.ds@gmail.com.
### Include both names of your tw-person group in the email.
### Deadline: 3:59 PM September 16, 2020

Write a function, that if given a character vector (string), returns the string, and then the number of characters in it. If more then one element is provided in the vector, return the results in a one row per element format.

Example:
```{r}
foo <- c(
    "I love data.table",
    "That sounded nerdy"
)

# countNumberOfCharacters(foo)
#> 'I love data.table' --> length: 17
#> 'That sounded nerdy' --> length: 18
```

Optional extra: Write a function, that if given two numeric vectors (`x` and `y`):
    - sums up the two vectors separately, and then multiplies the results together if the `sum_then_multiply` argument is `TRUE`
    - multiplies the two vectors element by element, and then sums up the results together if the `sum_then_multiply` argument is `FALSE`
    - `sum_then_multiply` should be `TRUE` by default