---
title: "Eltecon Data Science Course"
subtitle: "Introduction - Practice"
author: "Gabor Kocsis"
date: 'September 9, 2020'
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assigning objects
Store the sum of 24 and 42 in an object named `foo`, then divide `foo` by 3. What did you get as a result?

# Data types
### Vectors
```{r}
starks <- c("Arya", "Sansa", "Bran", "Rick")
```
Using the above defined `starks` vector, create a new character vector called `northeners` but extend it with "Cat" and "Ned".

Try out what happens when you mix different data types in a vector by using `class()` on the following vectors:
```{r}
nums_and_char <- c(1, 2, 3, 'a')
nums_and_logical <- c(1, 2, 3, TRUE)
chars_and_logical <- c('a', 'b', 'c', TRUE)
nums_and_char_again <- c(1, 2, 3, '4')
```

### Factors
```{r}
is_alive <- factor(c(TRUE, TRUE, TRUE, FALSE), levels = c(TRUE, FALSE), labels = c("alive", "dead"))
factor(is_alive)
```

Instead of typing in three times TRUE and two times FALSE, you can use the `rep(x, times)` function to replicate the value `x` the number of times set in the `times` parameter. Try to re-create `is_alive` uising `rep()`.

### Lists
```{r}
heights <- c(155, 175, 180, 145)
L <- list(A = starks, B = heights, C = is_alive)
L
```
In the list `L` - defined above - the second element contains the height of the Starks. Can you return it? Can you return just the height of Bran? Note that his height is the third element in the B sublist.

### Data frames
```{r}
GT <- data.frame(person = starks, height = heights, status = is_alive)
GT
```
Return the first and the third columns togehter and the second column separately by subsetting the `GT` data.frame. 

Use the `nrow()` function to get the number of rows in `GT`, then use this number to subset the data frame only to that row. 

Use the `summary()` function to calculate the mean and median height in the GT data frame. 

# Packages
Install the `data.table` package and then load it into R by using `library()`.

Once the `data.table` package is loaded in, check out the help of the Fast and friendly file finagler aka `fread` function.

# Functions
Write your own function which given a numeric vector, returns only those values from it that are above the average.

# Loops
Now modify the previous function so that instead of returning a vector, it prints out each value above the average in a separate line.
