---
title: "Data Viz - Class 1 - Notes"
author: "Peter Lukacs"
date: "9/22/2020"
output: html_document
---

```{r setup, include=FALSE}
local({
  r <- getOption("repos")
  r["CRAN"] <- "https://cran.rapporter.net//"
  options(repos = r)
})
```

## Install/load packages

```{r warning=FALSE, message=FALSE, results='hide'}
install.packages("ggplot2")
install.packages("data.table")
library(ggplot2)
library(data.table)
```

```{r}
theme_set(theme_minimal())
```

## mpg: mileage per gallon

```{r}
mpg <- unique(as.data.table(mpg))
head(mpg)
nrow(mpg)
```

## Simple plotting

> Always start with the data transformed to your needs
> then add the ggplot func. call with the data
> and finally add one or more layers

```{r}
ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy))
```

```{r}
mpg_agg <- mpg[, .(avg_displ = mean(displ)), by = drv]
ggplot(data = mpg_agg) +
    geom_col(mapping = aes(x = drv, y = avg_displ))
```

```{r}
mpg_agg <- mpg[, .(avg_cty = mean(cty)), year]
ggplot(data = mpg_agg) +
    geom_line(mapping = aes(x = year, y = avg_cty, group = 1))
```

## Adding more than just X and Y variables

```{r}
ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy, color = drv))
```

```{r}
# see what happens if you add the color param outside of the mapping function
ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy, color = drv), color = "blue")
```

```{r}
# the following wouldn't really make sense
ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy, size = drv))
```

```{r}
# use alpha to see through
ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy, size = drv), alpha = .3)
```

```{r}
# but this would sometimes (maybe not this time)
ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy, shape = drv), alpha = .3)
```

## Adding extra lines

```{r}
# ab line
ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy)) +
    geom_abline(slope = 1, intercept = 0, color = "red")
# see also hline/vline
```

```{r}
# linear regression line
ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy)) +
    geom_smooth(mapping = aes(x = cty, y = hwy), method = "lm")
```

## Facets

```{r}
ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy), alpha = .3) +
    facet_wrap(~drv)
```

```{r}
ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy), alpha = .3) +
    facet_wrap(~drv, ncol = 1)
```

```{r}
ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy, color = fl)) +
    facet_wrap(~drv, ncol = 1)
```

## Plot types where data aggregation happens while plotting 

### histogram

```{r}
ggplot(data = mpg) +
    geom_histogram(mapping = aes(x = hwy))
```

```{r}
ggplot(data = mpg) +
    geom_bar(mapping = aes(x = fl))
# Where does count come from?
```

```{r}
ggplot(data = mpg) +
    geom_bar(mapping = aes(x = fl, y = stat(prop), group = 1))
```

```{r}
ggplot(data = mpg) +
    geom_boxplot(mapping = aes(x = manufacturer, y = hwy))
```

## Beautifying the plot 

### labels

```{r}
ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy, color = class)) +
    labs(
        title = "Vehicle fuel consumption: city vs. highway",
        subtitle = "City vs. Highway",
        x = "City",
        y = "Highway",
        color = "Type of car",
        caption = "Fuel consumption is shown in milage per gallon (mpg)"
    ) +
    coord_fixed()
```

### coord(inate systems)

```{r}
# See how ugly the X axis labels are
ggplot(data = mpg) +
    geom_boxplot(mapping = aes(x = manufacturer, y = hwy))
```

```{r}
ggplot(data = mpg) +
    geom_boxplot(mapping = aes(x = manufacturer, y = hwy)) +
    coord_flip()
```

### zooming with coord

```{r}
ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy, color = class)) +
    coord_cartesian(xlim = c(15, 20), ylim = c(25, 30))
```

### scales: breaks and labels
```{r}
ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy, color = class)) +
    scale_y_continuous(breaks = seq(0, 45, by = 5)) +
    scale_x_continuous(labels = function(x) paste(x, "m/g"))
```

### saving your plot

```{r}
my_plot <- ggplot(data = mpg) +
    geom_point(mapping = aes(x = cty, y = hwy, color = class))
```{r}


```{r}
# Or you can skip the 'plot = ' part. (See the default for the 'plot' parameter!)
ggsave("figures/my_plot.png", plot = my_plot)
```
